---
title: "TEMPLATE - visualizing distributions"
author: "Your Name"
date: 2026-01-14
toc: true
---

::: {.callout-note}
This template follows the [visualizing distributions slides](https://eds-240-data-viz.github.io/course-materials/lecture-slides/SLIDES-distributions.html){target="_blank}. Please be sure to cross-reference the slides, which contain important information and additional context!
:::

## Setup

```{r setup}
#| eval: true
#| echo: true
#| message: false
#| warning: false
##~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
##                                    setup                                 ----
##~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~

#..........................load packages.........................
library(tidyverse)
library(chron)
library(naniar)
library(ggridges)
library(gghighlight)
library(ggbeeswarm)
library(palmerpenguins) # for some minimal examples

#..........................import data...........................
mko <- read_csv(here::here("week2", "data", "mohawk_mooring_mko_20250117.csv"))

##~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
##                                wrangle data                              ----
##~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~

mko_clean <- mko |>

  # keep only necessary columns ----
  select(year, month, day, decimal_time, Temp_bot, Temp_top, Temp_mid) |>

  # convert year, month, day & decimal_time cols to a date_time col ----
  # (not totally necessary for our plots, but helpful to know!)
  mutate(
  date_time = make_datetime(year, month, day, tz = "America/Los_Angeles") +
              seconds(decimal_time * 86400)  # decimal_time is fraction of day; 86400 seconds in a day
  ) |>

  # add month name by indexing the built-in `month.name` vector ----
  mutate(month_name = month.name[month]) |> 

  # replace 9999s with NAs ----
  naniar::replace_with_na(replace = list(Temp_bot = 9999, 
                                         Temp_top = 9999, 
                                         Temp_mid = 9999)) |>

  # select/reorder desired columns ----
  select(date_time, year, month, day, month_name, Temp_bot, Temp_mid, Temp_top)

##~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
##                            explore missing data                          ----
##~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~

# Missing data can have unexpected effects on your analyses
# It's important to explore your data for missing values so that you can decide if and how to handle them
# Data loggers can be prone to missingness (e.g. full memory, dead batteries, replacement logger)
# We can use {naniar} to explore the frequency and patterns in missing data
# Below is a short example of some {naniar} tools for doing so
# Check out The Missing Book (https://tmb.njtierney.com/) by Nick Tierney and Allison Horst for some great guidance

#..........counts & percentages of missing data by year..........
see_NAs <- mko_clean |> 
  group_by(year) |> 
  naniar::miss_var_summary() |>
  filter(variable == "Temp_bot")

#...................visualize missing Temp_bot...................
bottom <- mko_clean |> select(Temp_bot)
missing_temps <- naniar::vis_miss(bottom)
```

## Histograms & Density Plots

- great for visualizing the distribution of a numeric variable with one to a few groups

### Distinction

- **Histograms** show the counts (frequency) of values in each range (bin), represented by the height of the bars.
- **Density plots** show the relative proportion of values across the range of a variable. The total area under the curve equals 1, and peaks indicate values are more concentrated. Density plots do not show the absolute number of observations.

```{r dummy-hist-dens}
#| eval: true
#| echo: true
#| warning: false
#| message: false
#| layout-ncol: 2

#.....................create some dummy data.....................
dummy_data <- data.frame(value = c(rnorm(n = 100, mean = 5),
                                   rnorm(n = 200, mean = 10)),
                         group = rep(c("A", "B"),
                                     times = c(100, 200)))

#............................histogram...........................
ggplot(dummy_data, aes(x = value, fill = group)) +
  geom_histogram(position = "identity", alpha = 0.7) +
  geom_rug(aes(color = group), alpha = 0.75)

#..........................density plot..........................
ggplot(dummy_data, aes(x = value, fill = group)) +
  geom_density(alpha = 0.7) +
  geom_rug(aes(color = group), alpha = 0.75)
```

### Avoid plotting too many groups

```{r too-many-groups}
#| eval: true
#| echo: true
#| warning: false
#| message: false
#| layout-ncol: 2

#..................histogram with all 12 months..................
mko_clean %>% 
  mutate(month_name = factor(month_name, levels = month.name)) %>% 
  ggplot(aes(x = Temp_bot, fill = month_name)) +
  geom_histogram(position = "identity", alpha = 0.5)

#................density plot with all 12 months.................
mko_clean %>% 
  mutate(month_name = factor(x = month_name, levels = month.name)) %>%
  ggplot(aes(x = Temp_bot, fill = month_name)) +
  geom_density(alpha = 0.5)

```

### Consider faceting (small multiples)

```{r faceting}
#| eval: true
#| echo: true
#| warning: false
#| message: false
#| layout-ncol: 2

#...................histogram faceted by month...................
mko_clean |> 
  mutate(month_name = factor(month_name, levels = month.name)) |> 
  ggplot(aes(x = Temp_bot)) +
  geom_histogram() +
  facet_wrap(~month_name)

#..................density plot faceted by month.................
mko_clean |> 
  mutate(month_name = factor(month_name, levels = month.name)) |> 
  ggplot(aes(x = Temp_bot, fill = month_name)) +
  geom_density(fill = "gray30") +
  facet_wrap(~month_name)
```

### Consider plotting fewer groups + update colors + modify bin / band widths

- histograms & density plots are best suited for visualizing the distribution of one to a small number of groups
-  consider your scale of interest when deciding bin width, e.g. does a bin width of 1 (°C) make sense for our histogram? Does your density plot's bandwidth show the true underlying data distribution shape while reducing noise?

```{r fewer-groups}
#| eval: true
#| echo: true
#| warning: false
#| layout-ncol: 2

# hex codes for updating colors: "#2C5374", "#ADD8E6", "#8B3A3A"

#..................histogram with fewer months...................
mko_clean |> 
  mutate(month_name = factor(month_name, levels = month.name)) |> 
  filter(month_name %in% c("April", "June", "October")) %>% 
  ggplot(aes(x = Temp_bot, fill = month_name)) + 
  geom_histogram(position = "identity", alpha = 0.5, color = "black", binwidth = 1) +
  scale_fill_manual(values = c("#2C5374", "#ADD8E6", "#8B3A3A"))

#.................density plot with fewer months.................
mko_clean |> 
  filter(month_name %in% c("April", "June", "October")) %>%
  ggplot(aes(x = Temp_bot, fill = month_name)) +
  geom_density(alpha = 0.5, adjust = 0.5) +
  scale_fill_manual(values = c("#2C5374", "#ADD8E6", "#8B3A3A"))
```

::: {.callout-note}
## Question:
Why are the months in our density plot still in chronological order, despite not reordering them using `mutate()` (as we do for our histogram)?
:::

The month names are alphabetical so they don't need to be reordered to be in order

### Overlay histogram & density plots as a sanity check

::: {.callout-note}
## Question:
What should you carefully consider when checking the smoothing asumptions of your density curve against a histogram?
:::

The adjust argument will determine how much to smooth the line

```{r overlay-hist-dens}
#| eval: true
#| echo: true
#| warning: false
#| message: false

#..................overlay histogram & density...................
ggplot(mko_clean, aes(x = Temp_bot, y = after_stat(density))) +
  geom_histogram(fill = "gray", color = "black", alpha = 0.75) +
  geom_density(linewidth = 1)

```

## Ridgeline plots

- great for comparing distribution shapes of a numeric variable across many ordered groups 

```{r ridgeline}
#| eval: true
#| echo: true
#| warning: false
#| message: false
#| layout-ncol: 2

# hex codes for updating colors: "#2C5374","#849BB4", "#D9E7EC", "#EF8080", "#8B3A3A"

#......................basic ridgeline plot......................
mko_clean %>% 
  mutate(month_name = factor(month_name, levels = rev(month.name))) %>% 
  ggplot(aes(x = Temp_bot, y = month_name)) +
  ggridges::geom_density_ridges(rel_min_height = 0.01, scale = 3)

#....................fill with color gradient....................
mko_clean %>% 
  mutate(month_name = factor(month_name, levels = rev(month.name))) %>% 
  ggplot(aes(x = Temp_bot, y = month_name, fill = after_stat(x))) +
  ggridges::geom_density_ridges_gradient(rel_min_height = 0.01, scale = 3) +
  scale_fill_gradientn(colors = c("#2C5374","#849BB4", "#D9E7EC", "#EF8080", "#8B3A3A"))
```

## Boxplots

- great for *summarizing* the distribution of a numeric variable for multiple to many groups
- consider flipping your axes when you have long labels
- consider highlighting a group(s) of interest to focus attention

```{r basic-boxplot}
#| eval: true
#| echo: true
#| warning: false

#...................boxplot with all 12 months...................
ggplot(mko_clean, aes(x = month_name, y = Temp_bot, fill = month_name)) +
  geom_boxplot() +
  scale_x_discrete(limits = rev(month.name)) + # an alt way to reorder groups within ggplot, rather than during data wrangling
  gghighlight::gghighlight(month_name == "October") +
  coord_flip() +
  theme(legend.position = "none")

```

::: {.callout-note}
## Question:
What are the tradeoffs between reordering groups within a ggplot (as above) vs. during the data wrangling stage (e.g. as we did for our histogram and density and ridgeline plots)?
:::

### Overlay jittered data, if appropriate 

```{r boxplot-jitter}
#| eval: true
#| echo: true
#| warning: false

#........does not work if you have too many observations.........
ggplot(mko_clean, aes(x = month_name, y = Temp_bot)) +
  geom_boxplot(outlier.shape = NA) + # remember to remove outliers!
  geom_jitter(alpha = 0.5, width = 0.2) +
  scale_x_discrete(limits = rev(month.name)) + # an alt way to reorder groups within ggplot, rather than during data wrangling
  coord_flip()
#...........but great when you have fewer observations...........
ggplot(penguins, aes(x = species, y = body_mass_g)) +
  geom_boxplot(outlier.shape = NA) + # remember to remove outliers!
  geom_jitter(alpha = 0.5, width = 0.2) 

```

### Beeswarm plots as an alternative

```{r boxplot-beeswarm}
#| eval: true
#| echo: true
#| warning: false

##.......................boxplot + beeswarm.......................
ggplot(penguins, aes(x = species, y = body_mass_g)) +
  geom_boxplot(outlier.shape = NA) + # remember to remove outliers!
  ggbeeswarm::geom_beeswarm(size = 1)   
```

### Dodge when you have an additional grouping variable

```{r boxplot-dodge}
#| eval: true
#| echo: true
#| warning: false

#......................dodge groups by year......................
ggplot(penguins, aes(x = species, y = body_mass_g, color = factor(year))) +
  geom_boxplot(outlier.shape = NA) + # remember to remove outliers!
  geom_point(alpha = 0.5, 
             position = position_jitterdodge(jitter.width = 0.2))
```

## Violin plots

- great for visualizing the shape of the distribution of a numeric variable for multiple to many groups
- consider overlaying another chart type for added context

```{r violin-boxplot}
#| eval: true
#| echo: true
#| warning: false

#........................violin + box plot.......................
ggplot(mko_clean, aes(x = month_name, y = Temp_bot)) +
  geom_violin() +
  geom_boxplot(width = 0.1, fill = "gray", color = "gray30", alpha = 0.5) +
  scale_x_discrete(limits = rev(month.name)) +
  coord_flip()

```
